---
title: "Final"
author: "Felix Bridgeman & Liam Varela"
date: "2024-05-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(MASS)
library(biotools)
library(klaR)
library(car)
library(lubridate)
library(readxl)
library(ggExtra)
library(GGally)
library(plotly)
library(heplots)
library(nnet)
library(pROC)
library(corrplot)
library(aplpack)
library(fpc)
library(cluster)
library(ape)
library(vegan)
library(vegan3d)
library(mgcv)
library(rgl)
library(psych)
library(EFAtools)
library(knitr)
library(kableExtra)
library(dplyr)
library(pubtheme)
```

# Import Data

```{r}
dred <- read.csv("data/winequality-red.csv", sep = ";") %>%
  mutate(type = 1)
dwhite <- read.csv("data/winequality-white.csv" , sep = ";") %>%
  mutate(type = 2)

d <- rbind(dred, dwhite)

head(d)

# make ID column
d$ID <- 1:nrow(d)
# move ID column to front
d <- d[,c(ncol(d), 1:(ncol(d)-1))]

```

# Background and Motivation



# Research Question

- Which variables influence quality the most?
- Can we identify particular categories of wine based on combinations of variables?

# Data Description

We have 13 variables in our dataset.

| Variable | Type | Description |
|----------|------|-------------|
| ID | integer | unique identifier for each wine |
| type | categorical | red or white wine |
| quality | integer, categorical | rating of the wine from 0 to 10 with 10 being the best |
| fixed acidity | continuous | most acids involved with wine or fixed or nonvolatile |
| volatile acidity | continuous | the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste |
| citric acid | continuous | found in small quantities, citric acid can add 'freshness' and flavor to wines |
| residual sugar | continuous | the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet |
| chlorides | continuous | the amount of salt in the wine |
| free sulfur dioxide | continuous | the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine |
| total sulfur dioxide | continuous | amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine |
| density | continuous | the density of water is close to that of water depending on the percent alcohol and sugar content |
| pH | continuous | describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale |
| sulphates | continuous | a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant |
| alcohol | continuous | the percent alcohol content of the wine |

# Summaries of Data

```{r}
summary(d)
```

# Plots

```{r, fig.width=8, fig.height=10}
# plot histograms of all variables
dd <- d %>%
  pivot_longer(cols = -ID,
               names_to = "variable",
               values_to = "value")
g <- ggplot(dd,
            aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~variable, scales = "free",
             nrow = 5)

g

```

## Comments:

- pH appears normally distributed
- sulphates, alcohol, and quality appear right-skewed

## Now let's standardize data

```{r, fig.width=8, fig.height=10}
dc <- d
dc[-1] <- scale(d[-1])

dd <- dc %>%
  pivot_longer(cols = -ID,
               names_to = "variable",
               values_to = "value")

g <- ggplot(dd,
            aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~variable, scales = "free",
             nrow = 5)
g
```
















